---
title: "A cross-sectional study into the expression of financial challenges as expressed by ambulant hypertensive patients in Ghana"
author: "Author Information"
echo: false
warning: false
message: false
fontsize: "12"
bibliography: references.bib
citation: true
citations-hover: true
format: html
---

## Introduction

Hypertension or persistent high blood pressure is singularly responsible for most of the cardiovascular morbidities worldwide [@curado_direct_2021;@mahmood_association_2020;@mills_global_2020;@treciokiene_healthcare_2021] and almost a billion people in the world have hypertension @treciokiene_healthcare_2021. Age and a lower educational level have been identified as major risk factors for hypertension [@khan_prevalence_2021; @matei_relationship_2018]. There is also a strong bi-directional correlation between poverty and Non-Communicable Diseases (NCDs) [@WHO2017; @brummett_systolic_2019; @wang_association_2021] as NCDs pose serious threats to the financial stability of individuals and their families [@adane_cost_2020].

Direct medical cost is the biggest cost component for patients seeking medical care for hypertension @adane_cost_2020. Drug expenditure for hypertension is responsible for an increasing proportion of health costs and it accounts for 1.1 trillion United States Dollars (USD) annually worldwide [@athanasakis_socioeconomic_2017; @tamblyn_what_2018]. Among NCD sufferers, it has been documented that hypertensive patients particularly have a higher cost burden compared to non-hypertensive patients [@wang_association_2021;@zhang_research_2020;@valero-elizondo_favorable_2016]. The presence of a co-morbidity further exacerbates costs associated with the management of hypertension [@zhang_research_2020]. Among several factors, patient behaviors, such as adherence to treatment strongly affect healthcare expenditure due to hypertension [@petersen_strategies_2021]. Patients who had financial difficulties were also less likely to honor scheduled clinic visits [@michael_prevalence_2021].

In Australia, Out-of-pocket payment (OOP) alone is estimated to cost patients 941 million Australian dollars (AUS) per annum @adams_health_2020. A study in Columbia estimated the direct cost of hypertension care to cost households USD12,255.59 per annum [@londoñoagudelo2020]. In Ethiopia, So [@sorato_societal_2022] estimated the total direct cost to patients to be USD 64,837.48. In South African, an estimated USD 0.711 billion is OOP is spent annually by hypertensive patients [@kohli-lynch_hypertension_2022].

It is no surprise therefore that there have been several attempts worldwide to identify the ideal cost mechanisms to reduce the burden on hypertensive patients and the healthcare system in general [@jay_cost-effectiveness_2021; @wright2022; @treciokiene_healthcare_2021; @hannan_improving_2022; @oti_outcomes_2016]. The detection, treatment, and control of hypertension have been described as poor in Sub-Saharan Africa (SSA)[@cappuccio_cardiovascular_2016; @athanasakis_socioeconomic_2017; @mills_global_2020]. A global analysis of the situation by [@attaei_availability_2017] revealed huge disparities in the accessibility and affordability of HBP control medications. Residents in the upper-income countries had access and could afford 2 or more drug classes thereby having optimal clinical outcomes compared to residents in lower-income countries who had lesser access to BP medications and could not often afford their BPs medications. Hence, the blood pressure of residents in LMICs were often poorly controlled.

Hypertension is estimated to affect 13% of Ghanaians . Even though, the National Health Insurance Scheme (NHIS) provides financial cost protection for most people, particularly children and people with chronic diseases through the Ghana-Diagnostic Related Grouping (G-DRG) payment mechanism to hospitals, a good number of Ghanaians have still not subscribed [@sarkodie_effect_2021; @sanuade2018] owing principally to affordability [@kusi_refusal_2015] and a poor understanding of social health insurance [@kusi_refusal_2015; @kotoh_factors_2018]. Again, not all HBP medications are covered by the NHIS. In other instances too, the amount reimbursable is lower than the market price of the drugs [@harrison_anti-hypertensive_2021]. These factors often lead to catastrophic and impoverishing health expenditure for both subscribers and non-subscribers alike. The NHIS subscribers are not spared OOP while seeking health care. The NHIS has been shown to be inadequate in providing financial protections for subscribers [@de-graft_aikins_chronic_2014; @sarkodie_effect_2021] and OOP is still high among subscribers, sometimes paying fully [@sarkodie2021]. Frequent drug stock-outs [@sanuade_hypertension_2018] and delayed reimbursement to providers [@citifm_pay_2022; @agyepong2008] have resulted in situations where patients have to either buy their drugs from private pharmacies or make partial or full payments for services at the hospital [@sanuade_hypertension_2018-1].

Low-income households households are often not able to afford their medications and given that access and affordability to HBP medications is linked to clinical outcome, it is important the financial burden experienced by hypertensive patients is examined critically in Ghana and other LMICs. A study in Accra, Ghana, revealed 19.4% OOP rate among hypertensive patients while acquiring their drugs. As much as 40% could not afford their medication when they went to a private pharmacy to acquire their drugs and a further 25.3% could not afford their drugs at the hospital's pharmacy [@harrison_anti-hypertensive_2021]. This study examines the expression of financial difficulties or otherwise of hypertensive patients in 4 public hospitals in Kumasi and the patient characteristics associated with the experience of financial difficulties in accessing healthcare.

## Methodology

### Ethics

Ethical approval for the study was obtained from the Ghana Health Service and the ethical board of the Kwame Nkrumah University of Science and Technology. Prior approval was obtained separately from each hospital before data collection began. Data collection assistants were trained and supervised to collect data using a REDCap server [@Harris2009]. All patients were consented before data collection began.

### Patient recruitment

Patients that were attending routine medical care were recruited as part of the uHype study. The patients were asked as part of the questionnaire whether they experienced financial challenges accessing medical care including the cost drugs and laboratory services. Patient characteristics such as age and the presence of co-morbidities were recorded.

Data entry assistants were recruited and trained on the use of REDCap for data entry. The data entry assistants under the supervision of the team collected the data into the server.

### Statistical analysis

After the data collection, the data was exported to R where statistical analyses were performed. All packages used are cited in the references. The outcome variables for the study were derived from 3 related questions. One question elicited from patients whether they have financial difficulties attending the hospital for routine care. The other 2 questions asked whether there was financial difficulty in acquiring medications or laboratory services.

We built a logistic regression model with the outcome predictor being the expression of financial difficulties as a result of the HBP care. We used 450 patient records to build the regression model. We started with the null model and added variables and interactions until such a time the model Akaike Information Criterion (AIC) did not significantly change. Once a model was obtained, we used it to predict the outcome for the remaining 36 patients. These steps were repeated until the model successfully predicted the outcome for all 36 patients.

## Results and Analysis

### Characteristics of patients

```{r}
library(ggpubr)
library(gtsummary)
library(tidyverse)
uhype <- readRDS("uhype20230426.rds") %>%
  mutate(
    detained_12mths = fct_relevel(detained_12mths, "No", "Yes"),
    bp_meds___2 = fct_recode(
      bp_meds___2,
      "Co-payment" = "Yes",
      "NHIS covered" = "No"
    )
  )

theme_gtsummary_compact(font_size = 12)
theme_gtsummary_printer(print_engine = "flextable")
```

```{r}
dems <- uhype %>% 
  select(site, age, sex, educ, relationship, occup: no_occup, insurance: insurance_type___4, hpt_relative, mbp___1 : mbp___4) %>% 
  tbl_summary(
    missing = "no",
    type = list(
      everything() ~"categorical",
      age ~ "continuous"
    ),
    digits = everything() ~ c(0, 1)
  )
```

There were `r nrow(uhype)` patients recruited into the study. The study participants were recruited from 4 sites in the Kumasi metropolis (@tbl-dem). There were `r inline_text(dems, variable = site, level = 1)` study participants from the Ejisu Government Hospital, `r inline_text(dems, variable = site, level = 2)` from the Komfo Anokye Teaching hospital, and `r inline_text(dems, variable = site, level = 3)` from the KNUST hospital. The median (IQR) age of the study participants was `r inline_text(dems, variable = age)`. There were more females than males in the study, `r inline_text(dems, variable = sex, level = 1)`. About a 3^rd^, `r inline_text(dems, variable = educ, level = 1)`, of the participants had no education or did not complete primary school, almost half the of the participants, `r inline_text(dems, variable = educ, level = 2)` completed primary school or Junior High School, a further `r inline_text(dems, variable = educ, level = 4)` had post-secondary/high school education and `r inline_text(dems, variable = educ, level = 5)` had a tertiary education. Most, `r inline_text(dems, variable = relationship, level = "Partnership")`, of the patients had a partner (married or cohabiting). Of the `r inline_text(dems, variable = occup, level = "Yes")` that were engaged in an occupational activity, most of them were engaged in informal employment. For those who were not working `r inline_text(dems, variable = occup_type, level = 2)`, `r inline_text(dems, variable = no_occup, level = 2)` attributed the situation to ill health, `r inline_text(dems, variable = no_occup, level = 1)` were on retirement and `r inline_text(dems, variable = no_occup, level = 3)` had other non-specific reasons. Almost all of the patients, `r inline_text(dems, variable = insurance, level = 1)`, had insurance coverage, the majority [`r inline_text(dems, variable = insurance_type___1, level = 2)`] of whom were NHIS subscribers. Very few [`r inline_text(dems, variable = insurance_type___2, level = "Yes")`] of the patients had a private insurance scheme, of whom 2 were NHIS subscribers already. The only person that had an employee-paid program also had an NHIS subscription. Most [`r inline_text(dems, variable = hpt_relative, level = "Yes")`] of the patients knew at least one other family member that also had HBP, [`r inline_text(dems, variable = hpt_relative, level = 3)`]

Only `r inline_text(dems, variable = mbp___1, level = "Yes")` of the participants checked their BPs using a device they own or had access to in the house while a paltry `r inline_text(dems, variable = mbp___2, level = "Yes")` visited a neighborhood health facility to check their BPs and a further `r inline_text(dems, variable = mbp___3, level = "Yes")` visited pharmacy shops to check their BPs. Only `r inline_text(dems, variable = mbp___4, level = "Yes")` patients indicated they check their BPs with street vendors who carry BP apparatuses around to check for people.

```{r}
#| label: tbl-dem
#| tbl-cap: "Demographic characteristics of patients"

dems
```

### General risks and exposures

```{r}
rex <- uhype %>% 
  select(facilities_attended, more_than_1_facility, hpt_follow_up, med_hsp, same_doctor, hpt_special_clinic, med_runout, meds_concern,  dietician, special_diet, mea_bpm, bp_check_month, miss_appoint, health_rate, herbs_for_hpt, health_rate_after_1yr, comob, related, drugs, detained_12mths, feeling) %>% 
  tbl_summary(
    statistic = list(
      c(more_than_1_facility, facilities_attended) ~ "{median} (min = {min}, max = {max})"
    ),
    type = list(
      c(more_than_1_facility, facilities_attended) ~ "continuous"
    ),
    missing = "no"
  )
```

Most of the patients visited `r inline_text(rex, variable = facilities_attended)` facility for their healthcare needs of which `r inline_text(rex, variable = more_than_1_facility)` facility provided the patients with their BP medications (@tbl-rex). The median (IQR) travel time from their homes to the facility was `r inline_text(rex, variable = hpt_follow_up)` minutes. At the time of the visit, `r inline_text(rex, variable = med_hsp, level = "No")` had not taken their medications but `r inline_text(rex, variable = med_hsp, level = 1)` had taken all their medications before visiting the clinic. When asked how often they saw the same doctor, `r inline_text(rex, variable = same_doctor, level = "Never")` replied they never saw the same doctor, `r inline_text(rex, variable = same_doctor, level = "Sometimes")` said they saw the same doctor sometimes and only `r inline_text(rex, variable = same_doctor, level = "All times")` replied they saw the same doctor all the time. Most of the patients `r inline_text(rex, variable = hpt_special_clinic, level = "Yes")` said they visited special HPT clinics for their healthcare but `r inline_text(rex, variable = hpt_special_clinic, level = "I don't know")` were not aware whether they visited special clinics or not. Medications stock outs were infrequent among the patients: `r inline_text(rex, variable = med_runout, level = "Few times to never")` indicated they never run out of drugs and not more than 50 patients indicated they run out often or all the time. Few, `r inline_text(rex, variable = meds_concern)`, patients expressed concerns with the drugs. Not many  patients, `r inline_text(rex, variable = dietician)`, had consulted a dietitian for their BP control and a paltry `r inline_text(rex, variable = special_diet)` patients were on a special diet for BP control. Only `r inline_text(rex, variable = mea_bpm)` checked their BP in between visits and the median (IQR) number of times the patients checked their BP in between scheduled visits were `r inline_text(rex, variable = bp_check_month)`. Compliance with scheduled appointment was very high, `r inline_text(rex, variable = miss_appoint, level = 1)` and not more than 30 patients indicated some level of a high frequency of missing appointments. When the patients were asked about their perceived state of health, the majority of the patients indicated their health was between good to excellent with just about 20% indicating their health was either fair or poor. Notably, `r inline_text(rex, variable = herbs_for_hpt)` of the patients utilized herbal medications. On comparison of their current health with their health a year ago, about 200 of them expressed sentiments suggesting it was better. About a third of them suggested their perception of health was the same compared to a year ago. For about 70% of the patients, their perceived health was worse compared to the year before.

Only `r inline_text(rex, variable = comob, level = "None")` patients that not have an existing comorbidity, the majority, `r inline_text(rex, variable = comob, level = "One")` had 1 comorbidity at least. A further `r inline_text(rex, variable = comob, level = "Two")` of the patients had 2 co-morbidities and `r inline_text(rex, variable = comob, level = "More than 2")` had more than 2 comorbidities. Of the comobidities identified, `r inline_text(rex, variable = related)` of the patients had a comorbidity that has a high risk of a cardio-vascular accident in patients (diabetes, stroke, chronic kidney failure and heart failure). Most, `r inline_text(rex, variable = drugs, level = "More than 2")` the patients were on more than 2 medications for their hypertension management, `r inline_text(rex, variable = drugs, level = "Two")` were on exactly 2 medications and `r inline_text(rex, variable = drugs, level = "One")` were on a single medication for the BP control. Of the total number of patients interviewed, `r inline_text(rex, variable = detained_12mths)` had been admitted in a hospital within the last year on accounts of high blood pressure.

```{r #rexes}
#| label: tbl-rex
#| tbl-cap: General Financial risks and exposures

rex
```

### Financial exposure

```{r}
outs <- uhype %>% 
  select(ins_type___1:ins_type___2, miss_appoint, financial_difficulties, lab_test, bp_meds_buy, diff_get_drugs, med_runout,bp_meds___2, systol) %>% 
  tbl_summary(
    digits = everything() ~ c(0, 2),
    missing = "no",
    type = -systol ~ "categorical"
  )
```

Most, `r inline_text(outs, variable = ins_type___1, level = "Yes" )`, of the patients reported to the facility on the day of the interaction with their NHIS for healthcare but only `r inline_text(outs, variable = ins_type___2, level= "Yes")` had an additional private health insurance cover (@tbl-outcome). Most of the patients indicated they often comply with their review schedules, `r inline_text(outs, variable = miss_appoint, level = "Few times to never")`. Some of the patients `r inline_text(outs, variable = financial_difficulties, level = "Yes")` responded affirmative to having financial difficulties attending the health facility for healthcare. Most patients indicated they experienced financial difficulties accessing laboratory services, `r inline_text(outs, variable = lab_test, level = "Yes")`. Again, `r inline_text(outs, variable = bp_meds_buy, level = "Yes")` of the patients indicated they experienced challenges paying for their BP medications when they visited the facility of which `r inline_text(outs, variable = diff_get_drugs, level = "Less than half the time")` indicated they experienced the challenges less than half the time but `r inline_text(outs, variable = diff_get_drugs, level = "About half the time")` of the patients indicated they experienced the challenges about half the time, and `r inline_text(outs, variable = miss_appoint, level = "Sometimes to all the time")` patients indicated they experienced financial difficulties all the time paying for their drugs. Most of the patients `r inline_text(outs, variable = med_runout, level = "Few times to never")`, responded they run out of drugs few times to never while `r inline_text(outs, variable = med_runout, level = "Sometimes to all the time")` of the patients said the run out sometimes to all the time. On their current visit to the hospital. When asked, `r inline_text(outs, variable = bp_meds___2, level = 2)` patients made co-payments for their drugs. The median of the average systolic blood pressure of the patients (average calculated as resting BP taken 3 times 5 minutes apart) was `r inline_text(outs, variable = systol)`mm/Hg.

```{r}
#| label: tbl-outcome
#| tbl-cap: Risks for high financial exposure
#| include: true

outs
```


```{r}
#| include: false

library(lme4)
```

```{r}
#| include: false
#| cache: true
mod <- glm(
  formula =    financial_difficulties  ~ site + relationship + mbp___1 + mbp___3 + detained_12mths + bp_meds___2 + relationship * mbp___1,
  family = binomial(),
  data = uhype
)  %>%
  tbl_regression(
    intercept = F,
    exponentiate = T,
    add_estimate_to_reference_rows = T
  )

eff_mod <- glmer(
  formula =    financial_difficulties  ~ relationship*mbp___1 + mbp___1 + mbp___3 + detained_12mths + bp_meds___2 + (1 |site),
  data = uhype,
  family = binomial()
) %>% 
  tbl_regression(
    exponentiate = T,
    add_estimate_to_reference_rows = T
  )

mnMod <- tbl_merge(
  list(mod, eff_mod),
  tab_spanner = c("Regular GLM", "Randome-effect (Site)")
)
```

### Financial risk exposure for hypertensive patients

@tbl-fin-risk examines the relationship between some patient characteristics and the experience of financial difficulty, reported as Odds Ratios (OR), confidence interval and the probability of the estimates under the null hypothesis. The table displays 2 models: the first table shows the model with the site as a fixed effect, the second table shows the site as a random effect on the outcome. The site of care as a fixed effect was not statistically significant, however, participants who visited the university hospital were less likely to complain of financial difficulties compared to those that visited the Ejisu Government Hospital. The odds of complaining was higher in both the Teaching hospital and Kumasi South Hospital even though none of the estimates were statistically significant. Under both models, study participants with a partner had a statistically significant lower odds of experiencing financial difficulties compared to those that were single, `r inline_text(mnMod, variable = relationship, level = 2, pattern = "*FE*={estimate_1} (95% CI {conf.low_1}, {conf.high_1}; p={p.value_1}), _RE_={estimate_2} (95% CI {conf.low_2}, {conf.high_2}; p={p.value_2})")`. Participants with access to a BP machine at home, or those that checked their BP at nearby pharmacy shops all had a statistically significant lower odds complaining of financial difficulties, `r inline_text(mnMod, variable = 3, level = 2, pattern = "*FE*={estimate_1} (95% CI {conf.low_1}, {conf.high_1}; p={p.value_1}), _RE_={estimate_2} (95% CI {conf.low_2}, {conf.high_2}; p={p.value_2})")` and `r inline_text(mnMod, variable = 2, level = 2, pattern = "*FE*={estimate_1} (95% CI {conf.low_1}, {conf.high_1}; p={p.value_1}), _RE_={estimate_2} (95% CI {conf.low_2}, {conf.high_2}; p={p.value_2})")`, respectively, and being admitted in the last 12 months at any hospital was associated with an experience of financial difficulty by the participants, `r inline_text(mnMod, variable = 5, level = 2, pattern = "*FE*={estimate_1} (95% CI {conf.low_1}, {conf.high_1}; p={p.value_1}), _RE_={estimate_2} (95% CI {conf.low_2}, {conf.high_2}; p={p.value_2})")`. It was observed from analysis that patients who made co-payments for their drugs were 3 times more likely to complain of financial difficulty than those who care was covered completely by NHIS, `r inline_text(mnMod, variable = 6, level = 2, pattern = "*FE*={estimate_1} (95% CI {conf.low_1}, {conf.high_1}; p={p.value_1}), _RE_={estimate_2} (95% CI {conf.low_2}, {conf.high_2}; p={p.value_2})")`. Lastly, patients with a BP machine at home but in a partnership were more likely, `r inline_text(mnMod, variable = 7, level = 1, pattern = "*FE*={estimate_1} (95% CI {conf.low_1}, {conf.high_1}; p={p.value_1}), _RE_={estimate_2} (95% CI {conf.low_2}, {conf.high_2}; p={p.value_2})")`, to experience financial difficulties.

```{r}
#| label: tbl-fin-risk
#| include: true
#| tbl-cap: Experience of financial difficulties
mnMod 
```

```{r}
#| include: false
supp <- uhype %>% 
  tbl_cross(
    row = mbp___1,
    col = relationship
  ) %>% 
  add_p()

sysout <- rstatix::t_test(uhype, systol ~ financial_difficulties) %>% 
  pull(p)

meds <- table(uhype$financial_difficulties, uhype$bp_meds___2) %>% 
  chisq.test()
```

## Discussion

We have shown the probable factors associated with financial distress for hypertensive patients seeking care in hospitals in Ghana. The findings in this paper agrees with previous publications that explored the cost of care for hypertensive patients in Ghana and in other parts of the world. Per our model, visiting any of the hospitals did not affect the experience of financial distress by the study participants as the site of care was shown to be statistically not significant in the generalized linear model. Sites as a random effect on the other parameters also did not change the estimates by much. All the estimates tend to be close to each other, as such we can confidently say that the site of care has little to no effect on the cost of care for hypertensive patients. The presence of a partner as shown in our model significantly reduced financial distress, a phenomenon also observed in another study @bogale_pcost_2020. The phenomenon is explainable by the increased access to funds by the patient, something that will not be available to a single person. Access and availability to a blood monitoring device as shown in our model was also observed by [@krakoff2019] to be very cost effective besides it being as reliable as those checked in ambulatory settings. A study encourage health insurance providers to reimburse facilities that provide home monitoring @pickering2008. Just as home monitoring, contacts with a private pharmacies was shown to improve patient outcomes and even lower healthcare costs for hypertensive patients [@jay_cost-effectiveness_2021;@harrison_anti-hypertensive_2021]. The interaction of the having a partner and a BP monitoring device at home showed an interesting observation. Independently, they both reduced the financial difficulties by the patients but participant who had a device at home and were married were more likely to be financially distressed than their counterparts. The interaction is very interesting and we have not found any study that has explored the interaction between the two. A possible explanation is that the cost of the device may have exposed the household to a catastrophic expenditure. To further confirm our suspicions, we tested the association between having the device and being in a relationship but the test was not statistically significant, `r inline_text(supp, col_level = "p.value")`. We are very curious about this association and we would explore this in the future and we equally encourage other researchers to take this up.

From our model, we observed a disturbing pattern of people who are able to afford supplementary services are less likely to experience financial distress, given that they are already able to afford. Owning a home device or visiting pharmacy to check your BP involves some expenditure, something lower-income individuals/families may be unable to afford. This would imply those that are already financially distressed are further burdened by the cost of care compared to their colleagues. Despite this, there was no statistical association in the systolic BP of those that expressed financial distress and those that did not, p=`r sysout`. 

The NHIS, despite known challenges __insert references__ is still very important in offering financial protection for hypertensive patients in Ghana. Co-payments/NHIS as a predictor improved the model by reducing the Alkaike Information Criterion (AIC) from 607 to 574, making it a very important predictor of financial distress. This association was further confirmed by inspecting the residuals from the chi-square test between the financial distress and NHIS coverage. As can be seen in @fig-res, patients who cost were covered by the NHIS only were also more likely to complain of financial distress compared to those that paid partly for their drugs, further indicating those with lower incomes are at greater financial risk. We further checked our suspicion by checking the number of drugs each patient was taking for their hypertension. Patients with financial difficulties sometimes took fewer drugs fewer drugs than their counterparts, however, there was no statistical association once more @fig-trend.

```{r}
#| label: fig-res
#| fig-cap: Financial difficulty and Payment type chi-square residuals correlation plot 

meds$residuals %>% corrplot::corrplot(is.corr = F, addgrid.col = T, method = "pie")
```


```{r}
#| label: fig-trend
#| fig-cap: Relationship between financial distress and the number of drugs patients take

cor_txt <- rstatix::t_test(formula = multidrug ~ financial_difficulties, data = uhype, detailed = T) %>% 
  mutate(
    across(where(is.numeric), \(x) round(x, 2)),
    ntv = glue::glue("t(df),p={statistic}({df}),{p}")
  ) %>% 
  pull(ntv)


read_rds("fin_diff-n_drugs.rds") %>% 
  ggline(
  x =  "multidrug",
  y = "p",
  color = "financial_difficulties",
  ylab = "Percentage of patients",
  xlab = "Number of drugs",
  legend.title = "Financial difficulty",
  caption = cor_txt
)  + 
  scale_y_continuous(labels = scales::percent, expand = expansion(add = c(0, .05)), breaks = seq(0, .50, .05))  +
  ggsci::scale_color_lancet()
```


## Conclusion

## References
